Open Data, Open Source and Open Standards in chemistry: \\ The Blue Obelisk five years on

### 2011-06-03 paper.tex
Proofreading

The Quixote project is not dependent on any particular technology,
other than the representation of computational
chemistry in CML and the management of semantics through CML
dictionaries. At present, we use JUMBO-Converters for most
 of the semantic conversion, Lensfield2 for the workflow and Chempound
(chem\#) to store and disseminate the results.

@@

Open Source provides a unique opportunity for commercial organisations to work with the
cheminformatics community. Traditional business models rely on monetisation of source
code, causing every company to repeat the development by other companies, often combined
with a free (gratis) model for people working at academic institutes, to increase adoption
and allow contributions from academics. This solution defines the IP on the software
return on investment, but has the downside of investment losses due to duplication of
software and method development, which becomes visible when proprietary companies
merge. Some authors have rationalized that in the chemistry field, the
data has value, and few contributors are available to volunteer time
to improve codes.\cite{Stahl:2005fk} If true, this would hamper
adoption of Open Source and Open Data in chemistry, and greatly slow
the growth of projects such as those in the Blue Obelisk.

The Blue Obelisk community, however, takes advantage that much of the investments needed
for development are either payed by academic institutes and funding schemes, and by
volunteers investing time and effort who get full access to the source code in return.
In fact, all partners get full access up front and the Open Source licending ensures
that they will have access any time in the future. As such, it functions as a social
contract between everyone to arrange the immediate return on investment. Effectively,
this approach shares the burden of the high investment in having to develop cheminformatics
software from scratch, allowing researchers and commercial partners alike to focus
on their core business, rather than the development of prerequisites. As such, the rich
collection of Open Source cheminformatics tools provided by the Blue Obelisk
greatly reduces investment up front for new companies in the
cheminformatics market. Such advantages have also been noted in the
drug discovery field.\cite{DeLano:2005uq}\cite{Munos:2006vn}\cite{Geldenhuys:2006kx}

@@

For example, iChemLabs released the ChemDoodle Web Components library under the GPL v3
license, based on the upcoming HTML5 Open Standard. It allows making web and mobile
interfaces for chemical content. The project is already being adopted by others,
including iBabel by Chris Swain\cite{iBabel}, ChemSpotlight by
Geoffrey Hutchison\cite{chemspotlight} and the RSC ChemSpider\cite{chemspider_chemdoodle}.

>>>>

The Quixote project is not dependent on any particular technology,
other than the representation of computational
chemistry in CML and the management of semantics through CML
dictionaries. At present, we use JUMBO-Converters~\cite{JUMBO-Convertors} for most
 of the semantic conversion, Lensfield2~\cite{Lensfield2} for the workflow and Chempound
(chem\#) \cite{Chempound} to store and disseminate the results.

@@

Open Source provides a unique opportunity for commercial organisations to work with the
cheminformatics community. Traditional business models rely on monetisation of source
code, causing companies to repeat work done by other companies. This
model is sometimes combined
with a free (gratis) model for people working at academic institutes, to increase adoption
and encourage contributions from academics. This solution defines the IP on the software
return on investment, but has the downside of investment losses due to duplication of
software and method development, which becomes visible when proprietary companies
merge. Some authors have argued hat in the chemistry field
few contributors are available to volunteer time
to improve codes and IP considerations may prevent contributions from
industry~\cite{Stahl:2005fk}. If true, this would hamper
adoption of Open Source and Open Data in chemistry, and greatly slow
the growth of projects such as those in the Blue Obelisk.

The Blue Obelisk community, however, takes advantage of the fact that much of the investment needed
for development is either paid for by academic institutes and funding
schemes, or by
volunteers investing time and effort. In return, contributors get full
access to the source code, and the Open Source licensing ensures
that they will have access any time in the future. In this way, the
license functions as a social
contract between everyone to arrange the immediate return on investment. Effectively,
this approach shares the burden of the high investment in having to develop cheminformatics
software from scratch, allowing researchers and commercial partners alike to focus
on their core business, rather than the development of prerequisites.
In the case of the Blue Obelisk, the rich
collection of Open Source cheminformatics tools provided
greatly reduces investment up front for new companies in the
cheminformatics market. Such advantages have also been noted in the
drug discovery
field \cite{DeLano:2005uq}\cite{Munos:2006vn}\cite{Geldenhuys:2006kx}.

@@

For example, iChemLabs released the ChemDoodle Web Components library under the GPL v3
license, based on the upcoming HTML5 Open Standard. It allows making web and mobile
interfaces for chemical content. The project is already being adopted by others,
including iBabel~\cite{iBabel}, ChemSpotlight~\cite{chemspotlight} and the RSC ChemSpider~\cite{chemspider_chemdoodle}.
